Short-term neuronal and synaptic plasticity act in synergy for deviance detection in spiking networks

### 2023-03-20 paper-1/elife.tex
Better fig name: 5 -> thresholds follow std

\begin{figure}
    \includegraphics[width=\linewidth]{fig5}
    \caption{%
        Lower thresholds in \dev{} are a consequence of lower \dev{} non-target response magnitude.
        \textbf{A} $\T{A}{i}$ in the sample network, with each neuron shown in its spatial location. Left and middle column, raw averages in the \dev{} sequence ($\T[dev]{A}{i}$) and \msc{} sequence ($\T[msc]{A}{i}$), respectively; right column, contrast between these two ($\Delta \T{A}{i}$). A and B stimulus locations are highlighted in the left plot; see \FIG{1} for the locations of the remaining stimuli used in the \msc{} sequence.
        \textbf{B} Average response (in spikes per trial) to non-target trials ($\R{N}{i}$, with $N$ meaning B in \dev{}, and B, C, D, and E in \msc{}). Columns are arranged as in panel A.
        \textbf{C} Relationship across excitatory neurons of the sample network between $\Delta \R{N}{i}$ and $\Delta \T{A}{i}$, and associated linear regression.
        \textbf{D} $\Delta \R{N}{exc}$ across networks and stimuli, plotted against the Pearson correlation coefficients $\rho$ of the relationship shown in the previous panel ($\Delta \R{N}{i}$ vs $\Delta \T{A}{i}$ across excitatory neurons). Each marker corresponds to one network and target stimulus. Red markers indicate a significant positive correlation (one-sided Wald test, significance level 0.05). Inset histograms show the distribution of the corresponding values, with orange portions indicating the 95\% confidence interval of the median, calculated by bootstrap with 10000 iterations.
    }
    \label{fig:5}
    % 
    \figsupp[The equivalent figure in the full model with both TA and STD.]{%
        The equivalent of \FIG{5} in the full model with both TA and STD.
        \textbf{A} $\T{A}{i}$ in the sample network, with each neuron shown in its spatial location. Left and middle column, raw averages in the \dev{} sequence ($\T[dev]{A}{i}$) and \msc{} sequence ($\T[msc]{A}{i}$), respectively; right column, contrast between these two ($\Delta \T{A}{i}$). A and B stimulus locations are highlighted in the left plot; see \FIG{1} for the locations of the remaining stimuli used in the \msc{} sequence.
        \textbf{B} Average response (in spikes per trial) to non-target trials ($\R{N}{i}$, with $N$ meaning B in \dev{}, and B, C, D, and E in \msc{}). Columns are arranged as in panel A.
        \textbf{C} Relationship across excitatory neurons of the sample network between $\Delta \R{N}{i}$ and $\Delta \T{A}{i}$, and associated linear regression.
        \textbf{D} $\Delta \R{N}{exc}$ across networks and stimuli, plotted against the Pearson correlation coefficients $\rho$ of the relationship shown in the previous panel ($\Delta \R{N}{i}$ vs $\Delta \T{A}{i}$ across excitatory neurons). Each marker corresponds to one network and target stimulus. Red markers indicate a significant positive correlation (one-sided Wald test, significance level 0.05). Inset histograms show the distribution of the corresponding values, with orange portions indicating the 95\% confidence interval of the median, calculated by bootstrap with 10000 iterations.
    }
    {\includegraphics[width=\linewidth]{fig5-withSTD}}\label{figsupp:withSTD}
\end{figure}

First, we mapped the average value of $\theta$ at the start of target trials, $\T{A}{i}$, to the spatial location of the neurons, as shown in \FIG{5}A. In the \msc{} sequence, $\T[msc]{A}{i}$ was roughly evenly distributed across the network, consistent with the random, distributed nature of stimulation. Conversely, in the \dev{} sequence, a small number of neurons near the non-target stimulation site (B) were strongly adapted, whereas most of the network was less adapted, consistent with the data shown above.

Visualizing the likely cause of this difference, the non-target responses, in the same fashion (\FIG{5}B), we see that the pattern of activity in non-target trials, constituting 80\% of all trials, closely matched that of the thresholds, as expected. Notably, although the comparison is between 400 B trials in \dev{}, and only 100 B trials (alongside the same number of C, D and E trials) in \msc{}, only a small handful of neurons responded more to non-target stimulation in \dev{}, most of these very close to the stimulation site of B.

Correlating $\Delta \R{N}{i}$ with $\Delta \T{A}{i}$ across excitatory neurons (\FIG{5}C), we found that the reduced non-target response magnitude in \dev{} almost perfectly predicted the resulting reduction in TA levels before target stimuli (Pearson's $\rho$ = 0.99).
This finding held across networks and stimuli, as shown in \FIG{5}D, with a median correlation coefficient of 0.99 and a consistent reduction in the median non-target response $\R{N}{}$ between \msc{} and \dev{} sequences (T = 525, p = 2.12e-16, median = -0.094; one-sided Wilcoxon, n = 120), clearly validating our notion that the reduced non-target trial response was the primary driver of lower TA, which we showed above to be responsible for the higher target response.

@@

To show this, we sought to isolate this subset of early responders to B. In a given network and stimulus assignment, we ranked each neuron by the minimum response time (i.e., time to first spike) across B trials in all sequences. The ranking in the sample network is shown in \FIG{early-adaptation}A. We found that the response developed in an approximately radial pattern, spreading outward in all directions from the stimulus site. We did notice that the early response appeared to gravitate towards the stimulus site for A, but confirmed that this was the case in all sequences separately, indicating a structural rather than dynamic cause (data not shown). We then notionally split the network into an early portion, containing the first 100 neurons to respond to B, and a late portion, containing the remaining 900 neurons. In \FIG{early-adaptation}B, we related the non-target response contrast, $\Delta \R{N}{i}$, to the contrast in $\theta$ at the start of B trials, $\Delta \T{B}{i}$. Across excitatory neurons, a strong linear relationship (Pearson's $\rho$ = 0.993, n = 800) echoed the related finding in \FIG{5}C, where we correlated $\Delta \R{N}{i}$ to $\Delta \T{A}{i}$. As the color coding and inset histograms show, however, there was a clear separation between the early and late portions of the network, with many early neurons (red, left plot) increasing their response in the \dev{} sequence, while the response of the later portion of the network (blue, right plot) decreased. Correspondingly, the TA levels of the early portion also tended to increase, in clear contrast to the later portion.

Across networks and stimuli, we found even stronger evidence that the early portion of the network drove adaptation to B. We first confirmed that $\Delta \R{N}{i}$ and $\Delta \T{B}{i}$ were robustly positively correlated (median Pearson's $\rho$ 0.992: T = 7.26e+03, p = 9.86e-22; one-sided Wilcoxon, n = 120). In each network, we then identified the early and late portions with respect to stimulation in B and quantified the median non-target response contrast ($\Delta \R{N}{}$, \FIG{early-adaptation}C) and the median contrast in $\T{B}{}$ ($\Delta \T{B}{}$, \FIG{early-adaptation}D) within these two portions. We found that the response in the early portion was consistently greater in \dev{} than \msc{} ($\Delta \R{B}{early} > 0$: T = 6.98e+03, p = 8.48e-19), unlike the late portion, which decreased ($\Delta \R{B}{late} < 0$: T = 295, p = 1.23e-18; see also \FIG{5}D). $\Delta \T{B}{}$ followed the same pattern, with the early portion exhibiting stronger TA in \dev{} (T = 4.63e+03, p = 2.21e-17), while the late portion was instead more recovered (T = 26, p = 2.97e-17).

@@

\begin{figure}
    \includegraphics[width=\linewidth]{fig-addingSTD}
    \caption{%
        Adding STD to the model slightly enhances the effects of TA.
        \textbf{A} Reductions in mean response size $\mean{\R{A}{}}$ in each sequence as a consequence of adding STD to the model, across networks and stimuli.
        \textbf{B} $\Delta \R{N}{}$ (top) and $\Delta \T{B}{}$ (bottom) among early and late neurons, respectively, across networks and stimuli, in the full model with both STD and TA. Compare to \FIG{early-adaptation}C-D.
        \textbf{C} $\Delta \R{N}{exc}$ across networks and stimuli in the full model (horizontal axis) and the ablated model with TA only (vertical axis). The red line is the identity function; points above this line represent networks with greater response reduction (\dev{} - \msc{}) in the full model than in the ablated model. The subtraction of the two quantities in the box plot below shows that the non-target response tends to be reduced more in the full model.
        \textbf{D} $\Delta \R{N}{exc}$ across networks and stimuli in the full model, plotted against the corresponding Pearson coefficient of the correlation between $\Delta \R{N}{i}$ and $\Delta \T{A}{i}$ across excitatory neurons. Compare to \FIG{5}D.
        \textbf{E} $\Delta \T{A}{exc}$ across networks and stimuli in the full and ablated models, arranged as in panel C. The subtraction of the two quantities in the box plot below shows that $\T{A}{}$ tends to be reduced more in the full model.
        \textbf{F} $\Delta \T{A}{exc}$ across networks and stimuli in the full model, plotted against the corresponding Pearson coefficient of the correlation between $\Delta \T{A}{i}$ and $\Delta \R{A}{i}$ across excitatory neurons. Compare to \FIG{4}E.
        \textbf{G} $\Delta \R{A}{}$ across networks and stimuli in the full and ablated models, arranged as in panel C. The subtraction of the two quantities in the box plot below shows that the target response tends to be increased more in the full model.
    }
    \label{fig:addingSTD}
\end{figure}

@@

Comparing the non-target response reductions in \FIG{addingSTD}C, we found that they were on average larger in the full model than in the ablated model (T = 4.16e+03, p = 0.0175; one-sided Wilcoxon, n = 120), with the grand medians of $\Delta \R{N}{exc}$ differing by 36\%. Based on the ablated model analysis in \FIG{5}, we expected reduced non-target responses to similarly reduce the TA levels at the start of A trials. Indeed, $\Delta \R{N}{i}$ remained tightly correlated with $\Delta \T{A}{i}$ (median Pearson's $\rho$ = 0.989), as shown in \FIG{addingSTD}D. Correspondingly, $\T[dev]{A}{exc}$ was reduced over $\T[msc]{A}{exc}$, as shown by way of their contrast in \FIG{addingSTD}E, and this too by a slightly greater amount in the full model than in the ablated model (T = 4.37e+03, p = 0.0257).

>>>>

\begin{figure}
    \includegraphics[width=\linewidth]{fig-thresholds-follow-std}
    \caption{%
        Lower thresholds in \dev{} are a consequence of lower \dev{} non-target response magnitude.
        \textbf{A} $\T{A}{i}$ in the sample network, with each neuron shown in its spatial location. Left and middle column, raw averages in the \dev{} sequence ($\T[dev]{A}{i}$) and \msc{} sequence ($\T[msc]{A}{i}$), respectively; right column, contrast between these two ($\Delta \T{A}{i}$). A and B stimulus locations are highlighted in the left plot; see \FIG{1} for the locations of the remaining stimuli used in the \msc{} sequence.
        \textbf{B} Average response (in spikes per trial) to non-target trials ($\R{N}{i}$, with $N$ meaning B in \dev{}, and B, C, D, and E in \msc{}). Columns are arranged as in panel A.
        \textbf{C} Relationship across excitatory neurons of the sample network between $\Delta \R{N}{i}$ and $\Delta \T{A}{i}$, and associated linear regression.
        \textbf{D} $\Delta \R{N}{exc}$ across networks and stimuli, plotted against the Pearson correlation coefficients $\rho$ of the relationship shown in the previous panel ($\Delta \R{N}{i}$ vs $\Delta \T{A}{i}$ across excitatory neurons). Each marker corresponds to one network and target stimulus. Red markers indicate a significant positive correlation (one-sided Wald test, significance level 0.05). Inset histograms show the distribution of the corresponding values, with orange portions indicating the 95\% confidence interval of the median, calculated by bootstrap with 10000 iterations.
    }
    \label{fig:thresholds-follow-std}
    % 
    \figsupp[The equivalent figure in the full model with both TA and STD.]{%
        The equivalent of \FIG{thresholds-follow-std} in the full model with both TA and STD.
        \textbf{A} $\T{A}{i}$ in the sample network, with each neuron shown in its spatial location. Left and middle column, raw averages in the \dev{} sequence ($\T[dev]{A}{i}$) and \msc{} sequence ($\T[msc]{A}{i}$), respectively; right column, contrast between these two ($\Delta \T{A}{i}$). A and B stimulus locations are highlighted in the left plot; see \FIG{1} for the locations of the remaining stimuli used in the \msc{} sequence.
        \textbf{B} Average response (in spikes per trial) to non-target trials ($\R{N}{i}$, with $N$ meaning B in \dev{}, and B, C, D, and E in \msc{}). Columns are arranged as in panel A.
        \textbf{C} Relationship across excitatory neurons of the sample network between $\Delta \R{N}{i}$ and $\Delta \T{A}{i}$, and associated linear regression.
        \textbf{D} $\Delta \R{N}{exc}$ across networks and stimuli, plotted against the Pearson correlation coefficients $\rho$ of the relationship shown in the previous panel ($\Delta \R{N}{i}$ vs $\Delta \T{A}{i}$ across excitatory neurons). Each marker corresponds to one network and target stimulus. Red markers indicate a significant positive correlation (one-sided Wald test, significance level 0.05). Inset histograms show the distribution of the corresponding values, with orange portions indicating the 95\% confidence interval of the median, calculated by bootstrap with 10000 iterations.
    }
    {\includegraphics[width=\linewidth]{fig-thresholds-follow-std-withSTD}}\label{figsupp:withSTD}
\end{figure}

First, we mapped the average value of $\theta$ at the start of target trials, $\T{A}{i}$, to the spatial location of the neurons, as shown in \FIG{thresholds-follow-std}A. In the \msc{} sequence, $\T[msc]{A}{i}$ was roughly evenly distributed across the network, consistent with the random, distributed nature of stimulation. Conversely, in the \dev{} sequence, a small number of neurons near the non-target stimulation site (B) were strongly adapted, whereas most of the network was less adapted, consistent with the data shown above.

Visualizing the likely cause of this difference, the non-target responses, in the same fashion (\FIG{thresholds-follow-std}B), we see that the pattern of activity in non-target trials, constituting 80\% of all trials, closely matched that of the thresholds, as expected. Notably, although the comparison is between 400 B trials in \dev{}, and only 100 B trials (alongside the same number of C, D and E trials) in \msc{}, only a small handful of neurons responded more to non-target stimulation in \dev{}, most of these very close to the stimulation site of B.

Correlating $\Delta \R{N}{i}$ with $\Delta \T{A}{i}$ across excitatory neurons (\FIG{thresholds-follow-std}C), we found that the reduced non-target response magnitude in \dev{} almost perfectly predicted the resulting reduction in TA levels before target stimuli (Pearson's $\rho$ = 0.99).
This finding held across networks and stimuli, as shown in \FIG{thresholds-follow-std}D, with a median correlation coefficient of 0.99 and a consistent reduction in the median non-target response $\R{N}{}$ between \msc{} and \dev{} sequences (T = 525, p = 2.12e-16, median = -0.094; one-sided Wilcoxon, n = 120), clearly validating our notion that the reduced non-target trial response was the primary driver of lower TA, which we showed above to be responsible for the higher target response.

@@

To show this, we sought to isolate this subset of early responders to B. In a given network and stimulus assignment, we ranked each neuron by the minimum response time (i.e., time to first spike) across B trials in all sequences. The ranking in the sample network is shown in \FIG{early-adaptation}A. We found that the response developed in an approximately radial pattern, spreading outward in all directions from the stimulus site. We did notice that the early response appeared to gravitate towards the stimulus site for A, but confirmed that this was the case in all sequences separately, indicating a structural rather than dynamic cause (data not shown). We then notionally split the network into an early portion, containing the first 100 neurons to respond to B, and a late portion, containing the remaining 900 neurons. In \FIG{early-adaptation}B, we related the non-target response contrast, $\Delta \R{N}{i}$, to the contrast in $\theta$ at the start of B trials, $\Delta \T{B}{i}$. Across excitatory neurons, a strong linear relationship (Pearson's $\rho$ = 0.993, n = 800) echoed the related finding in \FIG{thresholds-follow-std}C, where we correlated $\Delta \R{N}{i}$ to $\Delta \T{A}{i}$. As the color coding and inset histograms show, however, there was a clear separation between the early and late portions of the network, with many early neurons (red, left plot) increasing their response in the \dev{} sequence, while the response of the later portion of the network (blue, right plot) decreased. Correspondingly, the TA levels of the early portion also tended to increase, in clear contrast to the later portion.

Across networks and stimuli, we found even stronger evidence that the early portion of the network drove adaptation to B. We first confirmed that $\Delta \R{N}{i}$ and $\Delta \T{B}{i}$ were robustly positively correlated (median Pearson's $\rho$ 0.992: T = 7.26e+03, p = 9.86e-22; one-sided Wilcoxon, n = 120). In each network, we then identified the early and late portions with respect to stimulation in B and quantified the median non-target response contrast ($\Delta \R{N}{}$, \FIG{early-adaptation}C) and the median contrast in $\T{B}{}$ ($\Delta \T{B}{}$, \FIG{early-adaptation}D) within these two portions. We found that the response in the early portion was consistently greater in \dev{} than \msc{} ($\Delta \R{B}{early} > 0$: T = 6.98e+03, p = 8.48e-19), unlike the late portion, which decreased ($\Delta \R{B}{late} < 0$: T = 295, p = 1.23e-18; see also \FIG{thresholds-follow-std}D). $\Delta \T{B}{}$ followed the same pattern, with the early portion exhibiting stronger TA in \dev{} (T = 4.63e+03, p = 2.21e-17), while the late portion was instead more recovered (T = 26, p = 2.97e-17).

@@

\begin{figure}
    \includegraphics[width=\linewidth]{fig-addingSTD}
    \caption{%
        Adding STD to the model slightly enhances the effects of TA.
        \textbf{A} Reductions in mean response size $\mean{\R{A}{}}$ in each sequence as a consequence of adding STD to the model, across networks and stimuli.
        \textbf{B} $\Delta \R{N}{}$ (top) and $\Delta \T{B}{}$ (bottom) among early and late neurons, respectively, across networks and stimuli, in the full model with both STD and TA. Compare to \FIG{early-adaptation}C-D.
        \textbf{C} $\Delta \R{N}{exc}$ across networks and stimuli in the full model (horizontal axis) and the ablated model with TA only (vertical axis). The red line is the identity function; points above this line represent networks with greater response reduction (\dev{} - \msc{}) in the full model than in the ablated model. The subtraction of the two quantities in the box plot below shows that the non-target response tends to be reduced more in the full model.
        \textbf{D} $\Delta \R{N}{exc}$ across networks and stimuli in the full model, plotted against the corresponding Pearson coefficient of the correlation between $\Delta \R{N}{i}$ and $\Delta \T{A}{i}$ across excitatory neurons. Compare to \FIG{thresholds-follow-std}D.
        \textbf{E} $\Delta \T{A}{exc}$ across networks and stimuli in the full and ablated models, arranged as in panel C. The subtraction of the two quantities in the box plot below shows that $\T{A}{}$ tends to be reduced more in the full model.
        \textbf{F} $\Delta \T{A}{exc}$ across networks and stimuli in the full model, plotted against the corresponding Pearson coefficient of the correlation between $\Delta \T{A}{i}$ and $\Delta \R{A}{i}$ across excitatory neurons. Compare to \FIG{4}E.
        \textbf{G} $\Delta \R{A}{}$ across networks and stimuli in the full and ablated models, arranged as in panel C. The subtraction of the two quantities in the box plot below shows that the target response tends to be increased more in the full model.
    }
    \label{fig:addingSTD}
\end{figure}

@@

Comparing the non-target response reductions in \FIG{addingSTD}C, we found that they were on average larger in the full model than in the ablated model (T = 4.16e+03, p = 0.0175; one-sided Wilcoxon, n = 120), with the grand medians of $\Delta \R{N}{exc}$ differing by 36\%. Based on the ablated model analysis in \FIG{thresholds-follow-std}, we expected reduced non-target responses to similarly reduce the TA levels at the start of A trials. Indeed, $\Delta \R{N}{i}$ remained tightly correlated with $\Delta \T{A}{i}$ (median Pearson's $\rho$ = 0.989), as shown in \FIG{addingSTD}D. Correspondingly, $\T[dev]{A}{exc}$ was reduced over $\T[msc]{A}{exc}$, as shown by way of their contrast in \FIG{addingSTD}E, and this too by a slightly greater amount in the full model than in the ablated model (T = 4.37e+03, p = 0.0257).
